```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error = FALSE, warning = FALSE)
```

## NTL Study Group Programming Task #1 - Health Data analysis
Code for this task will be demonstrated in both R & Python

### Data 
Data can be obtained from this [link](https://data.cdc.gov/NCHS/NCHS-Leading-Causes-of-Death-United-States/bi63-dtpu)

Data is downloaded and read from local harddisk

We start by loading needed libraries and reading the data
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(reticulate)

NHCS_data <- read.csv("NCHS.csv")
death_data <- read.csv("Death data.csv")
```

NHCS data file has ```r ncol(NHCS_data)``` columns & ```r nrow(NHCS_data)``` rows
Death data file has ```r ncol(death_data)``` columns & ```r nrow(death_data)``` rows

We start by checking how our data is structured
```{r}
str(NHCS_data)
str(death_data)
```

### Objectives
After checking structure of our data, we will use NHCS dataser to answer the few questions we have in hand .. 

1. What were the causes of death in this dataset?
```{r Q1}
Unique_causes <- unique(NHCS_data$Cause.Name)
Unique_causes[Unique_causes != "All Causes"]
```

2. What was the total number of deaths in the united states from 1999 to 2015?
```{r Q2}
# First we get a filtered version of our dataset
total_death_data <- filter(NHCS_data, NHCS_data$Year >= 1999 & NHCS_data$Year <= 2015, NHCS_data$Cause.Name == "All Causes", NHCS_data$State != "United States")

total_death <- sum(total_death_data$Deaths, na.rm = TRUE)
```
Total number of death is ```r total_death```

3. What is the number of deaths per year from 1999 to 2015?
```{r Q3}
# Sum Death per Year from 1999 to 2015
death_per_year <- aggregate(total_death_data['Deaths'], by=total_death_data['Year'], sum,na.rm=TRUE, na.action=NULL)

# Display death_per_year in a table
kable(death_per_year) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")

# Display Sum of Deaths per year from 1999 to 2015
ggplot(death_per_year, aes(Year, Deaths)) + geom_line()+labs(title = "Sum of Deaths per Year in the USA from 1999 to 2015")
```

4. Which states had the highest number of deaths overall?
```{r Q4}
# Sum Death per Year from 1999 to 2015
death_per_states <- aggregate(total_death_data['Deaths'], by=total_death_data['State'], sum,na.rm=TRUE, na.action=NULL)

# Display top 10 states with highest deaths in a table
kable(death_per_states[
    order( -death_per_states[,2]),
    ][1:10,]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")

# Display top 10 states with highest deaths
ggplot(death_per_states[
    order( -death_per_states[,2]),
    ][1:10,], aes(State, Deaths)) + geom_bar(stat = "identity", fill="blue", alpha=0.5)+labs(title = "Top 10 Deaths per State in the USA from 1999 to 2015") + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0))
```

5. What were top causes of death in the USA in this period?
```{r Q5}
# Sum Death per Cause from 1999 to 2015
total_USA_data <- filter(NHCS_data, NHCS_data$Year >= 1999 & NHCS_data$Year <= 2015, NHCS_data$Cause.Name != "All Causes", NHCS_data$State == "United States")

USA_data <- aggregate(total_USA_data['Deaths'], by=total_USA_data['Cause.Name'], sum,na.rm=TRUE, na.action=NULL)

# Display top 10 states with highest deaths in a table
kable(USA_data[
    order( -USA_data[,2]),
    ][1:10,]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")

# Display top 10 states with highest deaths
ggplot(USA_data[
    order( -USA_data[,2]),
    ][1:10,], aes(Cause.Name, Deaths)) + geom_bar(stat = "identity", fill="blue", alpha=0.5)+labs(title = "Top 10 Deaths per Cause in the USA from 1999 to 2015") + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0))
```